---
title: "Untitled"
author: "brouwern@gmail.com"
date: "July 18, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

### This script

* Certain years are typically referred to as "el nino" or "la nina" years, as oppossed to regualr years.  
* Climate indices can be used to assess the occurence and strength of these el nino/la nina years.  
* At http://ggweather.com/enso/oni.htm is a table by Jan Null (I believe) based on the Oceanic Niño Index (ONI) that ranks events as weak, strong, and very strong.
* I took the table and turned it into a spreadsheet
* This script processes this spreadsheet
* Since el ninos typically (always?) span from one calendar year to the next one year is the "leading" and the other "lagging" year (my terms)
* This script sets up the data to represent this


### For more information

* ONI data from NOA, with some color coding is here http://www.cpc.noaa.gov/products/analysis_monitoring/ensostuff/ensoyears.shtml
* This format is not easily downloaded
* Raw data appear to be here http://www.cpc.noaa.gov/products/analysis_monitoring/ensostuff/detrend.nino34.ascii.txt
* There is a package to download this stuff also https://github.com/boshek/rsoi/blob/master/README.md

### About ONI

From http://ggweather.com/enso/oni.htm
"The Oceanic Niño Index (ONI) has become the de-facto standard that NOAA uses for identifying El Niño (warm) and La Niña (cool) events in the tropical Pacific.  It is the running 3-month mean SST anomaly for the Niño 3.4 region (i.e., 5oN-5oS, 120o-170oW).  Events are defined as 5 consecutive overlapping 3-month periods at or above the +0.5o anomaly for warm (El Niño) events and at or below the -0.5 anomaly for cold (La Niña) events.  The threshold is further broken down into Weak (with a 0.5 to 0.9 SST anomaly), Moderate (1.0 to 1.4), Strong (1.5 to 1.9) and Very Strong (≥ 2.0) events.  For the purpose of this report for an event to be categorized as weak, moderate, strong or very strong it must have equaled or exceeded the threshold for at least 3 consecutive overlapping 3-month periods. [* Important note: This list was changed in July 2015 to reflect the use of ERSSTv4 by CPC in the calculation of Sea Surface Temperatures (SST)."



## Libraries
```{r}
library(ggplot2)
library(stringr)
```



## Load table of relative strengtsh
```{r}
nino.nom.str <- read.csv(file = "./data/data_in/el_nino_nominal_events_from_ONI_orig.csv",
                    skip = 3)
```


## Designate begining and ending year

Years in original table were concatenated, ie "1950-1951" because el ninos begin in the 2nd half of one year and resolve in the 1st half of the 2nd (I think)

```{r}
temp <- str_split(nino.nom.str$years,"-",simplify = T)


nino.nom.str$yr.strt <- as.numeric(temp[,1])

century <- gsub("([12][901])([01-9][01-9])","\\1", nino.nom.str$yr.strt )

nino.nom.str$yr.end <- as.numeric(paste(century,temp[,2], sep = ""))

```


## Fix years around millenia

The above gsub needs to be fixed for the year 2000
```{r}
i.millenia <- which(nino.nom.str$years == "1999-00")
nino.nom.str[i.millenia,"yr.end"] <- 2000
```

## Set up column names
```{r}
names(nino.nom.str) <- paste("nin",names(nino.nom.str),sep = ".")
names(nino.nom.str) <- gsub("year","yr",names(nino.nom.str) )
names(nino.nom.str) <- gsub("strength","str",names(nino.nom.str) )
names(nino.nom.str) <- gsub("start","strt",names(nino.nom.str) )

```

## Abbreviate words
```{r}
nino.nom.str$nin.str <- gsub("strong","str",nino.nom.str$nin.str)
nino.nom.str$nin.str <- gsub("weak","wk",nino.nom.str$nin.str)
nino.nom.str$nin.str <- gsub("very","vry",nino.nom.str$nin.str)
```




## Save

```{r}
write.csv(nino.nom.str, "./data/data_out/el_nino_nominal_events.csv", row.names = F)
```















Year a nino/a started
```{r}

nino.nom.str.starts <- nino.nom.str[,-dim(nino.nom.str)[2]]

names(nino.nom.str.starts)[1:3] <- paste(names(nino.nom.str.starts)[1:3],
                                    "strt",
                                    sep = ".")

names(nino.nom.str.starts)[4] <- "yr"


names(nino.nom.str.starts)


```

Year a nino/a ended
```{r}

nino.nom.str.ends <- nino.nom.str[,-4]

names(nino.nom.str.ends)[1:3] <- paste(names(nino.nom.str.ends)[1:3],
                                    "end",
                                    sep = ".")

names(nino.nom.str.ends)[4] <- "yr"


names(nino.nom.str.starts)


```







```{r}
dat6$nin.yrs.strt <- as.character(dat6$nin.yrs.strt)
dat6$nin.yrs.strt[is.na(dat6$nin.yrs.strt)] <- "no_event"
dat6$nin.yrs.strt <- factor(dat6$nin.yrs.strt)


dat6$nin.phase.strt <- as.character(dat6$nin.phase.strt)
dat6$nin.phase.strt[is.na(dat6$nin.phase.strt)] <- "no_event"
dat6$nin.phase.strt <- factor(dat6$nin.phase.strt)

dat6$nin.str.strt <- as.character(dat6$ nin.str.strt)
dat6$nin.str.strt[is.na(dat6$ nin.str.strt)] <- "no_event"
dat6$nin.str.strt <- factor(dat6$ nin.str.strt)


```

```{r}
dat6$nin.yrs.end <- as.character(dat6$nin.yrs.end)
dat6$nin.yrs.end[is.na(dat6$nin.yrs.end)] <- "no_event"
dat6$nin.yrs.end <- factor(dat6$nin.yrs.end)


dat6$nin.phase.end <- as.character(dat6$nin.phase.end)
dat6$nin.phase.end[is.na(dat6$nin.phase.end)] <- "no_event"
dat6$nin.phase.end <- factor(dat6$nin.phase.end)

dat6$nin.str.end <- as.character(dat6$ nin.str.end)
dat6$nin.str.end[is.na(dat6$ nin.str.end)] <- "no_event"
dat6$nin.str.end <- factor(dat6$ nin.str.end)

```



```{r}
dat6$nin.str.end <- factor(dat6$nin.str.end,
                    levels = c("vry.str",
                               "mod",
                               "wk",
                               "no_event"))

dat6$nin.str.strt <- factor(dat6$nin.str.strt,
                    levels = c("vry.str",
                               "mod",
                               "wk",
                               "no_event"))
```













